\textbf{The Illusion Game: A Novel Experimental Paradigm Provides Evidence in Favour of a General Factor of Visual Illusion Sensitivity and Personality Correlates}

### 2022-09-06 manuscript.Rmd
update manu

keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "1156"
bibliography      : "references.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

Notably, the presence of a common mechanism underlying the effect of different illusions has been contested [@hamburger2016; @cretenoud2019a; @teufel2018a; @cretenoud2020]; and the nature of the underlying processes - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences of prior beliefs [@teufel2018; @caporuscio2022] are strongly debated. The existence of dispositional or demographic correlates of illusion sensitivity - for example, higher illusion resistance has been reported in schizophrenia and autism [@notredame2014; @park2022; @giaouri2011; @keane2014]; as well as in males as compared to females [@miller2001; @papageorgiou2020; @lo2011; @shaqiri2018] (**Is there something for personality that we could use here instead of sex?**) - is another area of controversy.

@@

Study 3 aimed at investigating the links between the inter-individual scores of illusion sensitivity (obtained in study 2), and demographic and dispositional variables. **TODO: Insert a bit of literature about why and some findings justifying what we did**.

@@

The Bayesian t-tests on the effect of sex suggested anecdotal to moderate evidence in favour of the null effect for all scores, with the exception of the sensitivity to the Zöllner illusion, which was higher in males as compared to females ($\Delta=-0.37$, 95\% CI [-0.62, -0.13], $BF_{10} = 12.74$). We fitted Bayesian linear models with the education level entered as a monotonic predictor [appropriate for ordinal variables, @burkner2020modelling], which yielded no significant effects. For age, we fitted two types of models for each score, one general additive models (GAM) and a 2nd order polynomial model. These consistently suggested a significant positive linear relationship between age and Factor *i* ($pd=100\%$), as well as the sensitivity to Müller-Lyer ($pd=100\%$), Vertical-Horizontal ($pd=100\%$), Zöllner ($pd=100\%$) and Ebbinghaus ($pd=99\%$) illusions.

@@

Despite the widespread interest in finding associations between personality correlates and a general sensitivity to visual illusions, the supporting literature remains relatively mixed. Whereas some researchers report no relationship between inter-individual traits and individuals' susceptibility to illusions [@cretenoud2020a; @grzeczkowski2017], others have found pathological traits such as higher Aggression-Hostility and Narcissism, and lower Impulsive Sensation-Seeking to be significantly correlated to a greater resistance against such illusory effects [@zhang2017; @ohmann2016]. Furthermore, research on patients with schizophrenia have also generally demonstrated lower illusion sensitivity to be linked to symptoms of the disorder, such as greater hostility as well as social and emotional withdrawal [@pessoa2008].

>>>>

keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "1156"
bibliography      : "references.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

@@

Notably, the presence of a common mechanism underlying the effect of different illusions has been contested [@hamburger2016; @cretenoud2019a; @teufel2018a; @cretenoud2020]; and the nature of the underlying processes - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences of prior beliefs [@teufel2018; @caporuscio2022] are strongly debated. The existence of dispositional or demographic correlates of illusion sensitivity - for example, higher illusion resistance has been reported in schizophrenia and autism [@notredame2014; @park2022; @giaouri2011; @keane2014]; as well as for individuals with lower trait anxiety [@zhang2017; @qadri2020; @baroun2005] - is another area of controversy.

@@

Study 3 aimed at investigating the links between the inter-individual scores of illusion sensitivity (obtained in study 2), and demographic and dispositional variables. Despite the abundance of literature on visual illusions, relatively few researchers have attempted to examine the ties between demographic variables and personality traits with regard to individuals' resistance to illusions. Furthermore, most studies that focused on personality differences in illusion response primarily investigated the effects of traits associated with schizophrenia and autism spectrum disorders (such as impulsivity and sensation-seeking etc; [@hlavatá2018; @zhang2017; @lanyi2022can; @pessoa2008]). In this study, we instead aim to further elucidate inter-individual susceptibility to illusory effects by delineating the relationship between illusion susceptibility and demographic (i.e., age and gender) and dispositional components.

@@

The Bayesian t-tests on the effect of sex suggested anecdotal to moderate evidence in favour of the null effect for all scores, with the exception of the sensitivity to the Zöllner illusion, which was higher in males as compared to females ($\Delta=-0.37$, 95% CI [-0.62, -0.13], $BF_{10} = 12.74$). We fitted Bayesian linear models with the education level entered as a monotonic predictor [appropriate for ordinal variables, @burkner2020modelling], which yielded no significant effects. For age, we fitted two types of models for each score, one general additive models (GAM) and a 2nd order polynomial model. These consistently suggested a significant positive linear relationship between age and Factor *i* ($pd=100\%$), as well as the sensitivity to Müller-Lyer ($pd=100\%$), Vertical-Horizontal ($pd=100\%$), Zöllner ($pd=100\%$) and Ebbinghaus ($pd=99\%$) illusions.

@@

Interestingly, screen size was found to have a significant negative effect on illusion sensitivity for the White and Contrast illusions. This could be a result of our visual systems filtering through more low spatial frequency content when the size of the target object is small [@dixon2014]. As this filtering process excludes illumination information from visual processing, smaller screen sizes could result in an unprecedented change in brightness perception. As such, this could attenuate the illusory effect of the White and Contrast illusions, decreasing participants' susceptibility to the illusions since the illusory effects of both are dependent on luminance.

Although the present literature regarding the effects of personality correlates on illusion susceptibility remains fairly mixed [@cretenoud2020a; @grzeczkowski2017; @zhang2017; @ohmann2016], this study provides evidence of a positive relationship between illusion sensitivity and multiple personality traits, such as *Honesty-Humility* and *Agreeableness,* in healthy populations. Notably, higher scores in "pathological" traits such as *Antagonism, Psychoticism, Disinhibition* as well as *Negative Affect* were strongly correlated with lower illusion sensitivity (Factor *i*). This observation is consistent with past studies, in which higher self-reported Aggression-Hostility and Narcissism scores were associated with lower susceptibility to illusions [@zhang2017; @ohmann2016]. Moreover, low aggression and high self-control have also been found to correlate strongly with greater resistance to illusions [@mcgurk], conferring further support for the results we observed. As such, while the effects of personality on illusion sensitivity have typically only been found among clinical populations (usually in association with schizotypal traits, such as cognitive disorganization [@cretenoud2019]), this study demonstrates that differences in personality traits can also lead to inter-individual variation in illusion susceptibility. Further studies are required to replicate the observed findings and validate the underlying effects of personality dimensions on illusion susceptibility.
